---
title: "PS270 Final Project"
author: "Maeve Iliff"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to swing and battleground states. I am from Minnesota, which is the longest-standing "blue" state since 1976. But now I live in a key swing state- Wisconsin being a huge factor in presidential election victories. I want to look into data about what factors leads states to swing certain ways and how it changes across elections. I would like to look into candidate's platforms over the recent presidential elections, as well as geographic and demographic elements as well. Ultimately, the goal of the final project would be to get some sort of insight into what factors contribute to battleground states voting the way they do depending on the election year. 


## Research Proposal 

My research question is: How do the current state demographics- and the fluctuation between election years- relate to which president and party key battleground state constituents cast their vote to? My hypothesis is that changes in education levels, racial demographics, and geographic changes/movement determine which party is expected to win in a swing state. More education, diversity, and urbanization leads to Democratic victory.This is rooted in the modern realignment theory, which is the idea that shifts in voter allegiance shifts based on cultural and economic issues. My explanatory variables of interest are education levels (college graduates), percent/number of non-white voters in the state, and distribution of populations is urban and rural areas. I want to look at the variation and how they have changed in comparison to which party won the state's electoral college votes. My outcome variable will be the percentage of Democrat votes that given election year. To prove my hypothesis I would need to see positive change in regard to the variables in the same election cycle that the Democrat nominee won the presidency.  


## Code for Data to be Used in Project


I installed the tidycensus package to get the data for education, racial demography, and urbanization to compare across US censuses

```{r}
library(tidycensus)
```


Then I loaded data about voting by county to compare with my census data 

```{r}

library(tidyverse)

prescountydata <- read_csv("data/countypres_2000-2024.csv")

prescountydata

glimpse(prescountydata)

```



## Data Visualization for Milestone 3

I created a variable for all the swing states: Wisconsin, Michigan, Pennsylvania, Arizona, Georgia, Nevada, and North Carolina. Then I filtered into to data from swing states just in the year 2020. Lastly I filtered the 2020 swing state variable for democrat votes to be able to find the percentage of democrat votes in the total votes to use later in my data visualization. 

```{r}

swing_states <- c("WI", "MI", "PA", "AZ", "GA", "NV", "NC")

swing_2020 <- prescountydata |>
  filter(state_po %in% swing_states, year == 2020)

swing_2020 <- swing_2020 |>
  group_by(state_po, county_name) |>
  mutate(demvotes = sum(if_else(party == "DEMOCRAT", candidatevotes, 0))) |>
  mutate(demvoteshare = demvotes / totalvotes)

```

I looked into my tidycensus package to see all the avaliable functions within the package. I was able to use the get_acs() function to download data in each swing state county for people with a bachelor's degree (education) in 2020. Next, I matched county names to election data. Lastly, I needed to combine my swing_2020 variable data with the education data I just created and make sure everything was filtered by county for a more percise data visualization. 

```{r}

library(tidycensus)
ls("package:tidycensus")

acs_variables2020 <- load_variables(2020, "acs5")
head(acs_variables2020)

education <- get_acs(
  geography = "county", 
  state = swing_states, 
  variables = "B15003_022", 
  year = 2020)


education$county_name <- gsub(" County,.*", "", education$NAME)
education$county_name <- toupper(education$county_name)

swing2020_education <- merge(swing_2020, education, by = "county_name")


```



Lastly, I created a scatterplot to showcase my data to find a correlation between education and democratic votes in the counties of swing states. 


```{r}

swing2020_education_plot <- ggplot(swing2020_education, aes(x = estimate /1000, y = demvoteshare, color = state_po)) +
  geom_point(size = 1, alpha = 0.5) +
  geom_smooth(method = "lm", se= FALSE, color = "red") +
  scale_x_log10() +
  labs(
    x = "Number of Bachelor Degree Graduates in the Thousands", 
    y = "Democratic Vote Share", 
    title = "Education and Democratic Vote Share in Swing States (2020)", 
    color = "Swing State"
  )

swing2020_education_plot



```


## AI Disclosure Statement 

I used ChatGPT a couple times to create this data visualization. In my second chunk of code in focusing on the education variable, I was confused how to use the tidycensus library and all that code, so I inputed in to ChatGPT to ask to read it to find the names of data I wanted to use (for example that the average education is called estimate) because tidycensus takes data from the US census website so it is accissible that way. I looked up questions like "how to match my election data to the county data" and ChatGPT told me to use gsub and toupper, which I hadn't heard of before. I also asked it how to see the functions within the tidycensus library and it showed me the head() function and to use asc5 and to narrow down on 2020 and other variables specific to the census data. 





